\documentclass[11pt,a4paper]{article}

% ── Packages ──────────────────────────────────────────────────────────
\usepackage[margin=1in]{geometry}
\usepackage{xcolor}
\usepackage{booktabs}
\usepackage{longtable}
\usepackage{enumitem}
\usepackage{amssymb}
\usepackage{hyperref}
\usepackage{fancyhdr}
\usepackage{titlesec}
\usepackage{parskip}
\usepackage{graphicx}
\usepackage{array}
\usepackage{tikz}
\usepackage{tcolorbox}
\usepackage{multicol}

% ── Colors ────────────────────────────────────────────────────────────
\definecolor{googblue}{HTML}{4285F4}
\definecolor{googred}{HTML}{EA4335}
\definecolor{googgreen}{HTML}{34A853}
\definecolor{googyellow}{HTML}{FBBC05}
\definecolor{darkbg}{HTML}{1E1E2E}
\definecolor{lightgray}{HTML}{F5F5F5}
\definecolor{secgray}{HTML}{5F6368}
\definecolor{criticalred}{HTML}{D93025}
\definecolor{warningyellow}{HTML}{F9AB00}
\definecolor{okgreen}{HTML}{1E8E3E}

% ── Header / Footer ──────────────────────────────────────────────────
\pagestyle{fancy}
\fancyhf{}
\fancyhead[L]{\textcolor{secgray}{\small Google DeepMind — Engineering Audit}}
\fancyhead[R]{\textcolor{secgray}{\small CONFIDENTIAL}}
\fancyfoot[C]{\textcolor{secgray}{\small Page \thepage}}
\renewcommand{\headrulewidth}{0.4pt}

% ── Section styling ──────────────────────────────────────────────────
\titleformat{\section}{\Large\bfseries\color{googblue}}{}{0em}{}[\titlerule]
\titleformat{\subsection}{\large\bfseries\color{darkbg}}{}{0em}{}
\titleformat{\subsubsection}{\normalsize\bfseries\color{secgray}}{}{0em}{}

% ── Severity boxes ───────────────────────────────────────────────────
\newtcolorbox{criticalbox}{colback=criticalred!8,colframe=criticalred,title={\textbf{CRITICAL}},fonttitle=\color{white},coltitle=white}
\newtcolorbox{warningbox}{colback=warningyellow!10,colframe=warningyellow,title={\textbf{WARNING}},fonttitle=\color{black}}
\newtcolorbox{infobox}{colback=googblue!6,colframe=googblue,title={\textbf{NOTE}},fonttitle=\color{white},coltitle=white}
\newtcolorbox{strengthbox}{colback=okgreen!8,colframe=okgreen,title={\textbf{STRENGTH}},fonttitle=\color{white},coltitle=white}

\hypersetup{colorlinks=true,linkcolor=googblue,urlcolor=googblue}

% ══════════════════════════════════════════════════════════════════════
\begin{document}

% ── Title Page ───────────────────────────────────────────────────────
\begin{titlepage}
\centering
\vspace*{2cm}

{\Huge\bfseries\color{googblue} Engineering Audit Report}\\[0.6cm]
{\LARGE\color{darkbg} scry-chart v0.7.0}\\[0.4cm]
{\large\color{secgray} Pixel-Perfect TUI Charting Library for Rust}

\vspace{1.5cm}

\begin{tikzpicture}
  \fill[googblue] (0,0) rectangle (4,0.12);
  \fill[googred] (4.1,0) rectangle (8.1,0.12);
  \fill[googyellow] (8.2,0) rectangle (12.2,0.12);
  \fill[googgreen] (12.3,0) rectangle (16.3,0.12);
\end{tikzpicture}

\vspace{1.5cm}

\begin{tabular}{r l}
\textbf{Prepared by:} & Google DeepMind — Advanced Agentic Coding \\
\textbf{Date:}        & February 18, 2026 \\
\textbf{Repository:}  & \texttt{github.com/Slush97/scry} \\
\textbf{Crate:}       & \texttt{scry-chart} (within the scry monorepo) \\
\textbf{Classification:} & CONFIDENTIAL — Engineering Review \\
\end{tabular}

\vspace{1.5cm}

\begin{tcolorbox}[colback=lightgray,colframe=secgray,width=14cm]
\textbf{Executive Summary.}
\texttt{scry-chart} is a Rust charting library that renders anti-aliased, pixel-perfect
vector graphics directly in the terminal via the Kitty/Sixel protocols.
It ships 19 chart types, a 3D rendering pipeline, real-time streaming charts,
and export to PNG/SVG — all in ${\sim}12{,}000$ lines of Rust.

This audit evaluates its architecture, API design, visual quality, extensibility,
test coverage, and production readiness. Overall verdict:
\textbf{a strong, well-architected library with clear gaps in accessibility,
documentation depth, and test coverage that must be closed before a 1.0 release.}
\end{tcolorbox}

\vfill
{\small\color{secgray} Google DeepMind \textbullet\ Confidential Engineering Audit}
\end{titlepage}

% ── Table of Contents ────────────────────────────────────────────────
\tableofcontents
\newpage

% ═══════════════════════════════════════════════════════════════════════
\section{Scope \& Methodology}

\subsection{Scope}
Full source-code audit of \texttt{crates/scry-chart/} (v0.7.0), covering:
\begin{itemize}[nosep]
  \item Architecture \& module structure
  \item Public API surface \& ergonomics
  \item Visual rendering quality \& formatting
  \item Extensibility \& malleability
  \item Test coverage \& correctness
  \item Missing features vs.\ industry baselines (matplotlib, plotly, Vega-Lite)
  \item Security \& safety posture
\end{itemize}

\subsection{Methodology}
Static analysis of every \texttt{.rs} file in \texttt{src/}, outline-level review of
all public types, manual cross-referencing against \texttt{tests/} (71 test files)
and \texttt{examples/} (21 example files). No dynamic execution was performed
during this audit cycle.

% ═══════════════════════════════════════════════════════════════════════
\section{Architecture Overview}

\subsection{Module Dependency Graph}

\begin{center}
\begin{tikzpicture}[
  box/.style={draw,rounded corners,minimum height=0.8cm,minimum width=2.8cm,
              font=\small\ttfamily,fill=googblue!10},
  arr/.style={->,thick,color=secgray},
  node distance=1.2cm and 2.0cm
]
  \node[box] (chart) at (0,0) {chart/*};
  \node[box] (layout) at (4,0) {layout/*};
  \node[box] (scale) at (8,0) {scale};
  \node[box] (data) at (0,-1.8) {data};
  \node[box] (axis) at (4,-1.8) {axis};
  \node[box] (theme) at (8,-1.8) {theme};
  \node[box] (export) at (0,-3.6) {export};
  \node[box] (svg) at (4,-3.6) {svg\_export};
  \node[box] (widget) at (8,-3.6) {widget};
  \node[box,fill=googred!10] (engine) at (4,-5.2) {scry-engine};

  \draw[arr] (chart) -- (layout);
  \draw[arr] (layout) -- (scale);
  \draw[arr] (layout) -- (axis);
  \draw[arr] (axis) -- (scale);
  \draw[arr] (chart) -- (data);
  \draw[arr] (layout) -- (theme);
  \draw[arr] (export) -- (layout);
  \draw[arr] (svg) -- (layout);
  \draw[arr] (widget) -- (layout);
  \draw[arr] (export) -- (engine);
  \draw[arr] (widget) -- (engine);
  \draw[arr] (svg) -- (engine);
\end{tikzpicture}
\end{center}

\subsection{Layer Summary}

\begin{longtable}{>{\ttfamily}l p{9cm} r}
\toprule
\normalfont\textbf{Module} & \textbf{Responsibility} & \textbf{LoC (approx.)} \\
\midrule
chart/* & 19 chart type builders + enum dispatch & 3{,}800 \\
layout/* & Rendering pipeline: scales $\to$ axes $\to$ plot area $\to$ overlays & 4{,}200 \\
chart3d/* & 3D surface/scatter with camera, projection, wgpu backend & 4{,}500 \\
scale.rs & Linear, Log, Categorical, Time scales & 760 \\
axis.rs & Tick generation, label auto-skip, gridlines, zero-line & 930 \\
theme.rs & 6 built-in themes, hierarchical token system & 810 \\
legend.rs & 9-position legend with auto-best-corner & 535 \\
formatter/* & Auto, SI, Currency, Percent, Date, Locale, Null, Fn formatters & 1{,}400 \\
export.rs & PNG export, fontdue text stamping, subplot compositing & 530 \\
svg\_export.rs & SVG emit from scene graph, gradient defs, text overlays & 750 \\
data.rs & Series, SeriesStyle, FillPattern, GapPolicy & 560 \\
streaming.rs & Ring-buffered streaming chart with auto-scroll & 630 \\
widget.rs & Ratatui StatefulWidget, ChartState, interactivity & 540 \\
colormap.rs & 8 perceptual colormaps (Viridis, Plasma, RdBu, etc.) & 360 \\
Others & subplot, cursor, zoom, margin, annotation, inline, error & 950 \\
\midrule
\multicolumn{2}{l}{\textbf{Total source}} & $\mathbf{\sim12{,}300}$ \\
\bottomrule
\end{longtable}

\begin{strengthbox}
The codebase is remarkably compact for the feature surface it covers.
For comparison, \texttt{plotters} (Rust) is ${\sim}25{,}000$ LoC for comparable chart types
but lacks terminal pixel rendering, 3D, or streaming support.
\end{strengthbox}

% ═══════════════════════════════════════════════════════════════════════
\section{API Design \& Ergonomics}

\subsection{Builder Pattern}

Every chart type uses a fluent builder pattern terminating in \texttt{.build()}:

\begin{verbatim}
Chart::line(&data)
    .smooth()
    .filled()
    .title("Revenue")
    .theme(Theme::dark())
    .build()
\end{verbatim}

\begin{strengthbox}
The one-liner constructors (\texttt{Chart::line}, \texttt{Chart::scatter}, etc.)
are excellent for discoverability. The builder API surface is consistent across
all 19 chart types. Each builder returns a concrete type (e.g.\ \texttt{LineChart})
that implements type-specific methods, then \texttt{.build()} collapses into the
\texttt{Chart} enum. This is idiomatic Rust.
\end{strengthbox}

\begin{warningbox}
\textbf{Missing \texttt{TryBuild} / validation.}
\texttt{.build()} is infallible — invalid configurations (e.g.\ mismatched
\texttt{x}/\texttt{y} lengths, zero bins in histogram) silently produce degenerate output
instead of returning \texttt{Result}. This is acceptable for a plotting library (matplotlib
does the same), but a \texttt{.try\_build() -> Result<Chart, ChartError>} variant should
be offered for production pipelines.
\end{warningbox}

\subsection{Type Safety}

\begin{itemize}[nosep]
  \item \texttt{\#[must\_use]} on \texttt{Chart}, \texttt{ReferenceLine}, \texttt{SubplotGrid} — good.
  \item \texttt{\#[non\_exhaustive]} on all public enums and config structs — forward-compatible.
  \item \texttt{\#![deny(unsafe\_code)]} crate-wide — strong safety posture.
  \item \texttt{SeriesStyle} fields are all \texttt{Option<T>} with ``inherit from theme'' semantics — clean layering model.
\end{itemize}

\subsection{Prelude Completeness}

The prelude re-exports 30+ types. It is \textbf{well-curated} — covers the 90\%
use case without polluting the namespace. Notable exclusions that should be added:

\begin{itemize}[nosep]
  \item \texttt{FnFormatter} (the most common custom formatter path)
  \item \texttt{SubplotGrid} (currently requires explicit import)
  \item \texttt{StreamingChart} (already included — good)
\end{itemize}

% ═══════════════════════════════════════════════════════════════════════
\section{Visual Rendering Quality}

\subsection{Rendering Pipeline}

The pipeline follows a clean separation:

\begin{enumerate}[nosep]
  \item \textbf{Layout} — computes margins, axis positions, plot area rect
  \item \textbf{Scene graph} — emits \texttt{DrawCommand}s to \texttt{PixelCanvas} (from \texttt{scry-engine})
  \item \textbf{Rasterization} — \texttt{tiny-skia} renders anti-aliased primitives
  \item \textbf{Text stamping} — \texttt{fontdue} rasterizes Inter font glyphs
  \item \textbf{Transport} — Kitty/Sixel/halfblock protocol delivery or PNG/SVG file export
\end{enumerate}

\begin{strengthbox}
Using \texttt{tiny-skia} (a Skia subset in pure Rust) for rasterization is an excellent
choice — it provides sub-pixel anti-aliasing, gradient fills, and path rendering
without any C dependencies. The \texttt{fontdue} integration for text is similarly
zero-dependency and produces high-quality glyph output.
\end{strengthbox}

\subsection{Formatting \& Label Quality}

\subsubsection{Strengths}
\begin{itemize}[nosep]
  \item \textbf{Adaptive tick formatting} — \texttt{AutoFormatter} selects SI suffixes for
        large values, integer format for whole numbers, and decimal for small ranges.
        Uniform precision across all ticks on an axis via \texttt{format\_batch()}.
  \item \textbf{Auto-skip labels} — the \texttt{auto\_skip\_labels()} function drops every
        other label when they would overlap, considering rotation angle.
  \item \textbf{Nice domain rounding} — \texttt{LinearScale::nice()} pads and rounds domain
        bounds to clean numbers (handles degenerate/micro-range edge cases).
  \item \textbf{Locale support} — European number formatting (\texttt{1.234,56}).
  \item \textbf{Label rotation} — Horizontal, Diagonal (45°), Vertical (90°), or custom angle.
\end{itemize}

\subsubsection{Gaps}
\begin{warningbox}
\textbf{No automatic font-size scaling.} The font size is fixed (hardcoded at 11px
for regular text). Charts exported at different resolutions will have labels that
are proportionally too large or too small. A DPI-aware font-size heuristic is needed.
\end{warningbox}

\begin{warningbox}
\textbf{Title and axis labels lack smart wrapping.} Long titles or axis labels are
rendered as single-line text that can overflow the canvas. No word-wrap or
ellipsis truncation is implemented.
\end{warningbox}

% ═══════════════════════════════════════════════════════════════════════
\section{Feature Completeness}

\subsection{Chart Type Coverage}

\begin{longtable}{l c c c}
\toprule
\textbf{Chart Type} & \textbf{scry-chart} & \textbf{matplotlib} & \textbf{plotly} \\
\midrule
Line (smooth, step, filled) & \color{okgreen}\checkmark & \color{okgreen}\checkmark & \color{okgreen}\checkmark \\
Scatter & \color{okgreen}\checkmark & \color{okgreen}\checkmark & \color{okgreen}\checkmark \\
Bar (grouped, stacked) & \color{okgreen}\checkmark & \color{okgreen}\checkmark & \color{okgreen}\checkmark \\
Histogram & \color{okgreen}\checkmark & \color{okgreen}\checkmark & \color{okgreen}\checkmark \\
Box Plot & \color{okgreen}\checkmark & \color{okgreen}\checkmark & \color{okgreen}\checkmark \\
Violin Plot & \color{okgreen}\checkmark & \color{okgreen}\checkmark & \color{okgreen}\checkmark \\
Heatmap & \color{okgreen}\checkmark & \color{okgreen}\checkmark & \color{okgreen}\checkmark \\
Pie / Donut & \color{okgreen}\checkmark & \color{okgreen}\checkmark & \color{okgreen}\checkmark \\
Radar / Spider & \color{okgreen}\checkmark & \color{criticalred}$\times$ & \color{okgreen}\checkmark \\
Candlestick (OHLC) & \color{okgreen}\checkmark & \color{criticalred}$\times$ & \color{okgreen}\checkmark \\
Bubble & \color{okgreen}\checkmark & \color{okgreen}\checkmark & \color{okgreen}\checkmark \\
Waterfall & \color{okgreen}\checkmark & \color{criticalred}$\times$ & \color{okgreen}\checkmark \\
Funnel & \color{okgreen}\checkmark & \color{criticalred}$\times$ & \color{okgreen}\checkmark \\
Gauge / KPI & \color{okgreen}\checkmark & \color{criticalred}$\times$ & \color{okgreen}\checkmark \\
Lollipop & \color{okgreen}\checkmark & \color{criticalred}$\times$ & \color{criticalred}$\times$ \\
Sparkline & \color{okgreen}\checkmark & \color{criticalred}$\times$ & \color{okgreen}\checkmark \\
3D Surface & \color{okgreen}\checkmark & \color{okgreen}\checkmark & \color{okgreen}\checkmark \\
Streaming / Real-time & \color{okgreen}\checkmark & \color{warningyellow}$\sim$ & \color{okgreen}\checkmark \\
\midrule
Area Chart (standalone) & \color{criticalred}$\times$ & \color{okgreen}\checkmark & \color{okgreen}\checkmark \\
Treemap & \color{criticalred}$\times$ & \color{criticalred}$\times$ & \color{okgreen}\checkmark \\
Sankey Diagram & \color{criticalred}$\times$ & \color{criticalred}$\times$ & \color{okgreen}\checkmark \\
Contour Plot & \color{criticalred}$\times$ & \color{okgreen}\checkmark & \color{okgreen}\checkmark \\
Polar Plot & \color{criticalred}$\times$ & \color{okgreen}\checkmark & \color{okgreen}\checkmark \\
\bottomrule
\end{longtable}

\begin{infobox}
scry-chart covers 18/23 chart types compared against the union of matplotlib and plotly.
The missing types (Treemap, Sankey, Contour, Polar, standalone Area) are niche —
the current coverage is strong for a v0.7 release.
\end{infobox}

\subsection{Feature Matrix — Non-Chart Capabilities}

\begin{longtable}{l c l}
\toprule
\textbf{Feature} & \textbf{Status} & \textbf{Notes} \\
\midrule
PNG export & \color{okgreen}\checkmark & DPI-aware, fontdue text stamping \\
SVG export & \color{okgreen}\checkmark & Standalone SVG with gradient defs \\
PDF export & \color{criticalred}$\times$ & \textbf{Missing} — requires cairo or pdf-writer \\
Interactive (mouse) & \color{okgreen}\checkmark & Crosshair, tooltip, zoom, pan \\
Streaming / real-time & \color{okgreen}\checkmark & Ring buffer, auto-scroll \\
Subplot grid & \color{okgreen}\checkmark & Shared axes, gap control \\
Themes (6 built-in) & \color{okgreen}\checkmark & dark, light, pastel, ocean, forest, colorblind \\
Custom themes & \color{okgreen}\checkmark & Full token system \\
Colormaps (8) & \color{okgreen}\checkmark & Viridis, Plasma, Inferno, Magma, RdBu, etc. \\
Error bars & \color{okgreen}\checkmark & Symmetric ±error on Series \\
Annotations & \color{okgreen}\checkmark & Data-coordinate text labels \\
Reference lines & \color{okgreen}\checkmark & H/V lines with optional labels \\
Log scale & \color{okgreen}\checkmark & Base-10 with epsilon clamping \\
Time scale & \color{okgreen}\checkmark & Dedicated time axis formatter \\
Fill patterns & \color{okgreen}\checkmark & Hatched, crosshatched, dotted, diagonal \\
3D rendering & \color{okgreen}\checkmark & wgpu GPU backend + CPU fallback \\
Serde support & \color{okgreen}\checkmark & Optional feature flag \\
Inline (base64) & \color{okgreen}\checkmark & For notebook embedding \\
\midrule
Dual Y-axis & \color{criticalred}$\times$ & \textbf{Common request, not implemented} \\
Asymmetric error bars & \color{criticalred}$\times$ & Only symmetric \\
Animation / transitions & \color{criticalred}$\times$ & No keyframe/tween system \\
Data labels on points & \color{warningyellow}$\sim$ & Only via annotations (manual) \\
Responsive / auto-layout & \color{warningyellow}$\sim$ & Fixed px dims, no auto-resize \\
\bottomrule
\end{longtable}

% ═══════════════════════════════════════════════════════════════════════
\section{Extensibility \& Malleability}

\subsection{What's Easy to Extend}

\begin{itemize}[nosep]
  \item \textbf{New formatters} — implement the \texttt{TickFormatter} trait (3 methods) or
        use \texttt{FnFormatter} for quick closures. Clean trait-object erasure via
        \texttt{Arc<dyn TickFormatter>}.
  \item \textbf{New colormaps} — implement the \texttt{Colormap} trait (2 methods:
        \texttt{color\_at}, \texttt{name}).
  \item \textbf{New themes} — construct \texttt{Theme} with full token control
        (\texttt{TextStyle}, \texttt{AxisTheme}, \texttt{GridTheme}, \texttt{SeriesTheme}).
  \item \textbf{Per-series styling} — \texttt{SeriesStyle} allows surgical overrides
        without replacing the entire theme.
  \item \textbf{Custom export backends} — the scene graph (\texttt{PixelCanvas} + \texttt{DrawCommand}s)
        is public, so new output targets (PDF, Canvas, etc.) can walk the command list.
\end{itemize}

\subsection{What's Hard to Extend}

\begin{criticalbox}
\textbf{Adding new chart types requires modifying the \texttt{Chart} enum.}

The \texttt{Chart} type is a closed enum (\texttt{pub enum Chart \{ Scatter(...), Line(...), ... \}}).
Adding a 20th chart type requires:
\begin{enumerate}[nosep]
  \item Adding a variant to the enum
  \item Adding a match arm in every layout \texttt{match chart \{...\}} block
  \item Adding a constructor on \texttt{impl Chart}
\end{enumerate}
This is fine for the library maintainer but makes it \textbf{impossible for downstream
crates to define custom chart types}. A trait-based dispatch model (e.g.\
\texttt{trait ChartRenderer}) would be more extensible.
\end{criticalbox}

\begin{warningbox}
\textbf{Layout engine is monolithic.}
The layout module (\texttt{layout/mod.rs} + per-chart-type layout files) uses a single
large \texttt{render\_chart()} function with chart-type dispatch. Extracting the layout
pipeline into composable stages (margin $\to$ axes $\to$ plot $\to$ overlays) as
separate trait implementations would improve testability and extensibility.
\end{warningbox}

% ═══════════════════════════════════════════════════════════════════════
\section{Code Quality \& Formatting}

\subsection{Lint Configuration}

\begin{strengthbox}
Aggressive lint profile: \texttt{clippy::all}, \texttt{clippy::pedantic}, and
\texttt{clippy::nursery} are all set to \texttt{warn}. Combined with
\texttt{\#![deny(unsafe\_code)]} and \texttt{\#![warn(missing\_docs)]}, this is
one of the strictest lint configurations seen in the Rust ecosystem.
\end{strengthbox}

\subsection{Code Organization}

\begin{itemize}[nosep]
  \item \textbf{SPDX headers} on every file — good for license compliance.
  \item \textbf{Module doc comments} (\texttt{//!}) on every module — excellent.
  \item \textbf{Section separators} (ASCII rule lines) between logical blocks — consistent
        and readable.
  \item \textbf{Builder methods} documented with \texttt{///} doc comments including examples.
  \item \textbf{Separation of concerns}: chart definition (chart/*), layout rendering
        (layout/*), and output (export, svg\_export, widget) are cleanly separated.
\end{itemize}

\subsection{Issues}

\begin{warningbox}
\textbf{22 clippy \texttt{allow} overrides in \texttt{Cargo.toml} lints section.}
While some are justified (\texttt{cast\_precision\_loss}, \texttt{float\_cmp} for graphics code),
others suggest tech debt:
\begin{itemize}[nosep]
  \item \texttt{too\_many\_lines = "allow"} — some functions exceed 200 lines
  \item \texttt{indexing\_slicing = "allow"} — no bounds-check on slice indexing
  \item \texttt{many\_single\_char\_names = "allow"} — graphics math has many \texttt{x},
        \texttt{y}, \texttt{w}, \texttt{h} variables (acceptable but worth annotating)
\end{itemize}
\end{warningbox}

% ═══════════════════════════════════════════════════════════════════════
\section{Test Coverage}

\subsection{Test Infrastructure}

\begin{longtable}{l r l}
\toprule
\textbf{Category} & \textbf{Count} & \textbf{Notes} \\
\midrule
Integration test files & 71 & In \texttt{tests/} directory \\
Example files & 21 & In \texttt{examples/} directory \\
Inline unit tests & $\sim$40 & Scattered across modules \\
Snapshot tests (insta) & Yes & Glob-based via \texttt{insta} crate \\
Benchmarks & 1 & \texttt{chart3d\_bench} via Criterion \\
\bottomrule
\end{longtable}

\begin{warningbox}
\textbf{71 test files is impressive volume, but coverage is uneven.}
The SVG export module has 10 focused tests (structure, gradients, escaping).
However, the layout module — the most complex subsystem — does not appear to
have integration tests for edge cases like:
\begin{itemize}[nosep]
  \item Zero-length data series
  \item All-NaN data
  \item Single data point
  \item Extremely long labels (100+ characters)
  \item Very large datasets (100K+ points)
\end{itemize}
\end{warningbox}

\begin{criticalbox}
\textbf{No visual regression testing.}
For a charting library, pixel-level snapshot tests are essential. While \texttt{insta}
is configured, it appears to test serialized data structures rather than actual
rendered output. A render-and-diff pipeline (render PNG $\to$ compare against golden
reference) should be implemented.
\end{criticalbox}

% ═══════════════════════════════════════════════════════════════════════
\section{What They're Blatantly Missing}

\begin{longtable}{c l p{7.5cm}}
\toprule
\textbf{Priority} & \textbf{Gap} & \textbf{Impact} \\
\midrule
\textcolor{criticalred}{\textbf{P0}} & PDF export
  & Users exporting for reports cannot produce PDFs without external tooling.
    SVG-to-PDF is lossy. \\[4pt]

\textcolor{criticalred}{\textbf{P0}} & Visual regression tests
  & Any refactor risks silently breaking rendering. The library's core value
    proposition (``pixel-perfect'') is untestable. \\[4pt]

\textcolor{criticalred}{\textbf{P0}} & Dual Y-axis
  & Extremely common in dashboards. No workaround exists within the current API. \\[4pt]

\textcolor{warningyellow}{\textbf{P1}} & Responsive / auto-size
  & Charts require explicit pixel dimensions. Should auto-compute from terminal
    size or container. \\[4pt]

\textcolor{warningyellow}{\textbf{P1}} & Contour plot
  & Needed for scientific visualization. Pairs naturally with existing heatmap. \\[4pt]

\textcolor{warningyellow}{\textbf{P1}} & Font size scaling
  & Fixed 11px font is unreadable at small or large export sizes. \\[4pt]

\textcolor{warningyellow}{\textbf{P1}} & Asymmetric error bars
  & Only symmetric errors supported. Scientific data often has $(+a, -b)$. \\[4pt]

\textcolor{warningyellow}{\textbf{P1}} & Title/label wrapping
  & Long text silently overflows. Need word-wrap or ellipsis. \\[4pt]

\textcolor{googblue}{\textbf{P2}} & Data labels (auto)
  & Showing values on bars/points requires manual annotations. \\[4pt]

\textcolor{googblue}{\textbf{P2}} & Treemap / Sankey
  & Niche but expected by BI dashboard users. \\[4pt]

\textcolor{googblue}{\textbf{P2}} & Animation system
  & No transition/tween support for streaming chart updates. \\[4pt]

\textcolor{googblue}{\textbf{P2}} & Standalone area chart
  & Currently area fill is a line chart option. A dedicated \texttt{Chart::area()}
    constructor would improve discoverability. \\[4pt]

\textcolor{googblue}{\textbf{P2}} & Accessibility (WCAG)
  & Fill patterns exist (good) but no alt-text, ARIA, or screen-reader
    metadata in SVG output. \\
\bottomrule
\end{longtable}

% ═══════════════════════════════════════════════════════════════════════
\section{Security \& Safety}

\begin{strengthbox}
\textbf{Zero unsafe code.} The crate uses \texttt{\#![deny(unsafe\_code)]} and relies
entirely on safe Rust. No FFI, no raw pointer manipulation. All memory safety is
guaranteed by the compiler.
\end{strengthbox}

\begin{itemize}[nosep]
  \item \textbf{Dependency surface is minimal}: \texttt{tiny-skia}, \texttt{fontdue},
        \texttt{thiserror}, and optionally \texttt{ratatui}/\texttt{wgpu}. All are
        well-maintained crates.
  \item \textbf{No network access}: the library never makes HTTP requests or opens sockets.
  \item \textbf{Input validation}: scales handle degenerate domains (zero-span, NaN).
        Tick generation has a hard cap of 200 ticks to prevent infinite loops.
  \item \textbf{SVG output}: \texttt{xml\_escape()} sanitizes text content against XSS
        when generating SVG (10 unit tests verify this).
\end{itemize}

\begin{warningbox}
\textbf{Potential DoS vectors:}
\begin{itemize}[nosep]
  \item Extremely large data series (millions of points) with no automatic decimation
        or level-of-detail. A single \texttt{Chart::scatter} with 10M points will
        allocate proportional memory for the scene graph.
  \item \texttt{SubplotGrid} allows arbitrary dimensions — a $1000 \times 1000$ grid
        would attempt to allocate 1M cells.
\end{itemize}
These are low-risk for a charting library but worth documenting.
\end{warningbox}

% ═══════════════════════════════════════════════════════════════════════
\section{Comparative Analysis}

\subsection{vs.\ Rust Ecosystem}

\begin{longtable}{l c c c c}
\toprule
\textbf{Capability} & \textbf{scry-chart} & \textbf{plotters} & \textbf{ratatui-charts} & \textbf{textplots} \\
\midrule
Anti-aliased rendering & \color{okgreen}\checkmark & \color{okgreen}\checkmark & \color{criticalred}$\times$ & \color{criticalred}$\times$ \\
Terminal-native & \color{okgreen}\checkmark & \color{criticalred}$\times$ & \color{okgreen}\checkmark & \color{okgreen}\checkmark \\
Chart type count & 19 & 12 & 3 & 2 \\
3D support & \color{okgreen}\checkmark & \color{criticalred}$\times$ & \color{criticalred}$\times$ & \color{criticalred}$\times$ \\
GPU acceleration & \color{okgreen}\checkmark & \color{criticalred}$\times$ & \color{criticalred}$\times$ & \color{criticalred}$\times$ \\
Streaming / real-time & \color{okgreen}\checkmark & \color{criticalred}$\times$ & \color{criticalred}$\times$ & \color{criticalred}$\times$ \\
Interactive (mouse) & \color{okgreen}\checkmark & \color{criticalred}$\times$ & \color{criticalred}$\times$ & \color{criticalred}$\times$ \\
SVG export & \color{okgreen}\checkmark & \color{okgreen}\checkmark & \color{criticalred}$\times$ & \color{criticalred}$\times$ \\
Themes & 6 & 2 & 0 & 0 \\
\texttt{unsafe}-free & \color{okgreen}\checkmark & \color{criticalred}$\times$ & \color{okgreen}\checkmark & \color{okgreen}\checkmark \\
\bottomrule
\end{longtable}

\begin{strengthbox}
\texttt{scry-chart} is the most feature-complete terminal charting library in the
Rust ecosystem by a significant margin. Its combination of pixel-perfect terminal
rendering, 3D support, GPU acceleration, and interactivity is unique — no other
Rust crate offers this combination.
\end{strengthbox}

% ═══════════════════════════════════════════════════════════════════════
\section{Recommendations}

\subsection{Pre-1.0 Blockers (P0)}

\begin{enumerate}[nosep]
  \item \textbf{Visual regression testing} — Implement a render-to-PNG golden-file
        comparison pipeline. Block releases on pixel-diff thresholds.
  \item \textbf{PDF export} — Add \texttt{pdf-writer} or \texttt{printpdf} integration
        as an optional feature flag.
  \item \textbf{Dual Y-axis support} — Add \texttt{.y2\_axis()} to line/scatter builders
        with a second scale and axis on the right side.
  \item \textbf{Font size scaling} — Compute font size as a function of
        \texttt{min(width, height)} and DPI.
\end{enumerate}

\subsection{High-Value Improvements (P1)}

\begin{enumerate}[nosep]
  \item \textbf{Trait-based chart dispatch} — Extract \texttt{trait ChartType} to allow
        downstream chart implementations without modifying the enum.
  \item \textbf{Auto data labels} — \texttt{.show\_values()} on bar/histogram/lollipop builders.
  \item \textbf{Smart text wrapping} — Wrap long titles/labels with line breaks and ellipsis.
  \item \textbf{Level-of-detail decimation} — Auto-downsample large datasets for render performance.
  \item \textbf{Asymmetric error bars} — \texttt{.with\_error\_asymmetric(low, high)}.
  \item \textbf{Contour plot} — Natural extension of existing heatmap infrastructure.
\end{enumerate}

\subsection{Nice-to-Have (P2)}

\begin{enumerate}[nosep]
  \item SVG accessibility metadata (ARIA roles, \texttt{<title>}, \texttt{<desc>})
  \item Standalone \texttt{Chart::area()} constructor
  \item Animation/transition framework for streaming updates
  \item \texttt{try\_build()} validation variant
  \item Documentation cookbook with real-world examples
\end{enumerate}

% ═══════════════════════════════════════════════════════════════════════
\section{Overall Assessment}

\begin{center}
\begin{tabular}{l c}
\toprule
\textbf{Dimension} & \textbf{Score (1--10)} \\
\midrule
Architecture \& Design & \textbf{9} / 10 \\
API Ergonomics & \textbf{8} / 10 \\
Feature Completeness & \textbf{8} / 10 \\
Visual Quality & \textbf{8} / 10 \\
Extensibility & \textbf{6} / 10 \\
Test Coverage & \textbf{6} / 10 \\
Documentation & \textbf{7} / 10 \\
Security / Safety & \textbf{9} / 10 \\
Production Readiness & \textbf{7} / 10 \\
\midrule
\textbf{Overall} & \textbf{7.6} / 10 \\
\bottomrule
\end{tabular}
\end{center}

\begin{infobox}
\textbf{Bottom line:} \texttt{scry-chart} is an impressive, well-engineered charting
library that punches well above its weight at ${\sim}12{,}000$ LoC. It uniquely combines
pixel-perfect terminal rendering with a feature set that rivals desktop plotting libraries.
The architecture is clean, the API is ergonomic, and the safety posture is exemplary.

The primary risks are: (1) the closed \texttt{Chart} enum limits third-party extensibility,
(2) the absence of visual regression tests means refactors carry rendering risk, and
(3) several common features (dual Y-axis, PDF export, font scaling) are missing.

\textbf{Recommendation: Strong foundation for a 1.0 release after addressing P0 items.}
\end{infobox}

\end{document}
